# Recalibrate Legacy Dates

Further information regarding the radiocarbon dates discussed below can be found in the text of the previous article [@RN21001]. A table of associated diagnostic artifacts is included as a means of providing information related to relative dates.

```{r install, echo = TRUE}
devtools::install_github('ISAAKiel/oxcAAR')
# load oxcAAR and OxCal
library(kableExtra)
library(oxcAAR)
quickSetupOxcal()
```

## Calibration and plots for dates from Burial Pit 5 at the Mounds Plantation site

Three dates (Tx-55, Tx-56, and M-1466) were reported from Burial Pit 5 at the Mounds Plantation site. Additional information about these dates can be found in the text, and those works cited within the text of the manuscript.

```{r mplcal, echo=TRUE}
moundspl<-data.frame(bp=c(860, 900),
                     std=c(120, 100),
                     names=c("Tx-55", "M-1466")
)
moundsplcal<-oxcalCalibrate(moundspl$bp, 
                            moundspl$std, 
                            moundspl$names)
moundsplcal
plot(moundsplcal)
calcurve_plot(moundsplcal)
```

## Calibration and plots for dates from Burial Pit 2 at the Gahagan Mound site

Three AMS dates (UGA12296, ISGS A0465, and ISGS A0466) were reported from Burial Pit 2 at the Gahagan Mound site. Additional information about these dates can be found in the text, and those works cited within the text of the manuscript.

```{r gahagancal, echo=TRUE}
gahagan<-data.frame(bp=c(1000, 950, 910),
                    std=c(40, 40, 35),
                    names=c("UGA-12296", "ISGS-A0466", "ISGS-A0465")
)
gahagancal<-oxcalCalibrate(gahagan$bp, 
                           gahagan$std, 
                           gahagan$names)
gahagancal
plot(gahagancal)
calcurve_plot(gahagancal)
```

## Calibration and plots for the dates from F119 and F134 at the George C. Davis site

Two dates (Tx-913 and Tx-1206) were reported from Features 119 and 134, respectively. Additional information about these dates can be found in the text and cited references.

```{r GCDcal, echo=TRUE}
gcd<-data.frame(bp=c(1150, 986),
                std=c(80, 90),
                names=c("Tx-913", "Tx-1206")
)
gcdcal<-oxcalCalibrate(gcd$bp,
                       gcd$std,
                       gcd$names)
gcdcal
plot(gcdcal)
calcurve_plot(gcdcal)
```
