# Composition & Diversity

This analysis of assemblage diversity is limited to Caddo mortuary contexts at the Mounds Plantation, George C. Davis, and Gahagan Mound sites where Gahagan bifaces were recovered.

## Taxonomic composition

```{r composistion}
library(here)
library(tidyverse)

# read data
data <- read.csv("gahagan-diversity.csv")

# table of diagnostics
knitr::kable(data)
```

## Alpha diversity

```{r diversity, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
library(vegan)

# assemblage size (N)
N <- rowSums(data[11:25])
N ## assemblage sizes range from:

# how many of each type were found?
T <- colSums(data[11:25])
T

## richness (S) = number of types in assemblage
S <- specnumber(data[11:25])
S

## ubiquity (U) = number of assemblages that contain a particular type
U <- specnumber(data[11:25])
U
```

## Relative abundance

```{r relative.abundance}
# mean number of observations for each type
colMeans(data[11:25])

# data by percentage
data.pct <- data[11:25]/N*100
## mean percent (Mp) of each type across assemblage
Mp <- colMeans(data.pct)
Mp

## percentage of sites that have each type
Up <- U/length(N)*100
Up
```

### Relative abundance by context

```{r plot.burial.diversity, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap="Contexts ordered temporally from bottom to top based upon seriation results provided in the previous chapter. The middle two contexts (16RR1-BP2 and 16CD12-BP5) are interchangeable."}
library(dplyr)
library(reshape)
library(ggplot2)
library(ggpubr)
library(ggExtra)
library(wesanderson)

# relative abundance of juvenile and adult burials
pal <- wes_palette("Moonrise2", 2, type = "continuous")

burials <- data %>% 
  select(context, total_adult, total_juvenile) %>% 
  mutate(context = paste(context, c("(S)", "(S)", "(S)", "(S)",
                                    "(N)", "(N)", "(N)", "(N)"))) %>% 
  melt(id.vars = "context")

# configure plot
plot0 <- burials %>%
  arrange(context) %>% 
  mutate(context = factor(context, levels = c(
                            "16CD12-BP1 (N)", "41CE19-F134 (S)", "16RR1-BP3 (S)",
                            "16CD12-BP5 (N)", "16RR1-BP2 (S)", "41CE19-F119 (S)", 
                            "16CD12-BP2 (N)", "16CD12-BP8 (N)"))) %>% 
  ggplot(aes(x=context, y = value, fill = variable)) + 
  geom_bar(stat="identity", position = "fill") + 
  coord_flip() +
  scale_fill_manual(values = pal) +
  labs(x = "Context",
       y = "Relative abundance (%)",
       fill = "Category")

# relative abundance of burials by category
pal <- wes_palette("Moonrise2", 8, type = "continuous")

burials <- data %>% 
  select(context, adult_male:uid_adult, juvenile_male:uid_juvenile) %>% 
  mutate(context = paste(context, c("(S)", "(S)", "(S)", "(S)",
                                    "(N)", "(N)", "(N)", "(N)"))) %>% 
  melt(id.vars = "context")

# configure plot
plot00 <- burials %>%
  arrange(context) %>% 
  mutate(context = factor(context, levels = c(
                            "16CD12-BP1 (N)", "41CE19-F134 (S)", "16RR1-BP3 (S)",
                            "16CD12-BP5 (N)", "16RR1-BP2 (S)", "41CE19-F119 (S)", 
                            "16CD12-BP2 (N)", "16CD12-BP8 (N)"))) %>% 
  ggplot(aes(x=context, y = value, fill = variable)) + 
  geom_bar(stat="identity", position = "fill") + 
  coord_flip() +
  scale_fill_manual(values = pal) +
  labs(x = "Context",
       y = "Relative abundance (%)",
       fill = "Category")

# render figure
figure <- ggarrange(plot0, plot00,
                      labels = c("a","b"),
                      ncol = 1, nrow = 2)

# plot figure
figure
```

```{r plot.alpha.diversity, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap="Contexts ordered temporally from bottom to top based upon seriation results provided in the previous chapter. The middle two contexts (16RR1-BP2 and 16CD12-BP5) are interchangeable."}
# relative abundance of lithics + ceramics
pal <- wes_palette("Moonrise2")

cerlith <- data %>% 
  select(context, lithics, ceramics) %>% 
  mutate(context = paste(context, c("(S)", "(S)", "(S)", "(S)",
                                    "(N)", "(N)", "(N)", "(N)"))) %>% 
  melt(id.vars = "context")

# configure plot
plot1 <- cerlith %>%
  arrange(context) %>% 
  mutate(context = factor(context, levels = c(
                            "16CD12-BP1 (N)", "41CE19-F134 (S)", "16RR1-BP3 (S)",
                            "16CD12-BP5 (N)", "16RR1-BP2 (S)", "41CE19-F119 (S)", 
                            "16CD12-BP2 (N)", "16CD12-BP8 (N)"))) %>% 
  ggplot(aes(x=context, y = value, fill = variable)) + 
  geom_bar(stat="identity", position = "fill") + 
  coord_flip() +
  scale_fill_manual(values = pal) +
  labs(x = "Context",
       y = "Relative abundance (%)",
       fill = "Category")

# relative abundance of diagnostic types
pal <- wes_palette("Moonrise2", 17, type = "continuous")

diagnostics <- data %>% 
  select(context, alba:ceramic_bowl) %>% 
  mutate(context = paste(context, c("(S)", "(S)", "(S)", "(S)",
                                    "(N)", "(N)", "(N)", "(N)"))) %>% 
  melt(id.vars = "context")

# configure plot
plot2 <- diagnostics %>% 
  arrange(context) %>% 
  mutate(context = factor(context, levels = c(
                            "16CD12-BP1 (N)", "41CE19-F134 (S)", "16RR1-BP3 (S)",
                            "16CD12-BP5 (N)", "16RR1-BP2 (S)", "41CE19-F119 (S)", 
                            "16CD12-BP2 (N)", "16CD12-BP8 (N)"))) %>% 
  ggplot( aes(x=context, y = value, fill = variable)) + 
  geom_bar(stat="identity", position = "fill") + 
  coord_flip() +
  scale_fill_manual(values = pal) +
  labs(x = "Context",
       y = "Relative abundance (%)",
       fill = "Types") +
  theme(legend.key.height = unit(0.05, "cm"))

# render figure
figure <- ggarrange(plot1, plot2,
                      labels = c("a","b"),
                      ncol = 1, nrow = 2)

# plot figure
figure
```

## Diversity indices

```{r diversity.indices, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# diversity - diagnostics
# Simpson diversity
data$D1 <- diversity(data[11:25], index = "simpson")
D1 <- data$D1

# evenness - diagnostics
# Shannon diversity
data$H <- diversity(data[11:25])
H <- data$H

# Pielou's J (Shannon diversity index divided by natural log of richness)
data$J <- H/log(S)
J <- data$J
```

## Beta diversity - diagnostics

```{r diversity2, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap="Caddo burial assemblages that include Gahagan bifaces; 1, 41CE19-F134; 2, 41CE19-F119; 3, 16RR1-BP2; 4, 16RR1-BP3; 5, 16CD12-BP1; 6, 16CD12-BP2; 7, 16CD12-BP5; 8, 16CD12-BP8. The three sites in the lower part of the left quadrat exhibit lower diversity and evenness, and each represents the earliest context at each site to include a Gahagan biface. Those sites in the upper right quadrat have higher diversity and evenness, and occur later than those in the lower left quadrat."}
# summarize assemblage diversity to identify high & low diversity assemblages
library(maptools)
pch <- c(1, 3)[as.factor(data$region)]
plot(D1, J, 
     pch = pch)
abline(h = median(J), 
       v = median(D1), 
       lty = 2)
pointLabel(D1, J, 
           rownames(data), 
           cex = .75)
leg.txt <- c(as.expression(bquote("Northern Behavioral Region")), 
             as.expression(bquote("Southern Behavioral Region")))
legend("bottomright", 
       leg.txt, 
       pch = c(1, 3))
```

### Diversity through time

The figure above shows higher diversity and evenness in later contexts. Is there a significant increase in diversity between the initial burial context at each site that included Gahagan bifaces, and those that occur later?

```{r simpson.init.diversity, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# diversity
# mutate data to add temporal assignment
temp <- data %>%
  mutate(temporal = c(1, 2, 2, 1, 1, 2, 2, 2), .after = region)

# table of diagnostics
knitr::kable(temp, "pipe")

# attributes for plots
temp.D1 <- temp$D1
temp.J <- temp$J
time <- temp$temporal

#diversity
## plot diversity
ggplot(temp, aes(x = time, y = temp.D1)) +
  geom_point(aes(shape = factor(context))) +
  scale_shape_manual(values = c(0:5,15,17)) +
  geom_smooth(method = "lm")

# are data not normally distributed?
shapiro.test(temp.D1)
hist(temp.D1, col='steelblue')

# MODEL: D1 as a function of initial v. later contexts
temp.simpson <- aov(temp.D1 ~ time, 
                    data = temp)

# ANOVA: do temporal contexts differ by D1?
summary(temp.simpson)
```

### Evenness through time

The figure above shows higher diversity and evenness in later contexts. Is there a significant increase in evenness between the initial burial context at each site that included Gahagan bifaces, and those that occur later?

```{r simpson.init.evenness, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# evenness

## plot evenness
ggplot(temp, aes(x = time, y = temp.J)) +
  geom_point(aes(shape = factor(context))) +
  scale_shape_manual(values = c(0:5,15,17)) +
  geom_smooth(method = "lm")

# are data not normally distributed?
shapiro.test(temp.J)
hist(temp.J, col='steelblue')

# MODEL: J as a function of initial v. later contexts
temporal.J <- anosim(temp.J, 
                     grouping = time, 
                     permutations = 10000)

# ANOSIM: do temporal contexts differ by J?
summary(temporal.J)
```
