# Composition & Diversity

This analysis of assemblage diversity is limited to Caddo mortuary contexts at the Mounds Plantation, George C. Davis, and Gahagan Mound sites where Gahagan bifaces were recovered.

## Taxonomic composition

```{r composistion}
library(here)
library(dplyr)
# read data
data <- read.csv("gahagan-diagnostics.csv")

# table of diagnostics
knitr::kable(data)
```

## Alpha diversity

```{r diversity, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
library(vegan)

# assemblage size (N)
N <- rowSums(data[3:15])
N ## assemblage sizes range from:

# how many of each type were found?
T <- colSums(data[3:15])
T

## richness (S) = number of types in assemblage
S <- specnumber(data[3:15])
S

## ubiquity (U) = number of assemblages that contain a particular type
U <- specnumber(data[3:15])
U
```

## Relative abundance

```{r relative.abundance}
# mean number of observations for each type
colMeans(data[3:15])

# data by percentage
data.pct <- data[3:15]/N*100
## mean percent (Mp) of each type across assemblage
Mp <- colMeans(data.pct)
Mp

## percentage of sites that have each type
Up <- U/length(N)*100
Up
```

### Relative abundance by context

```{r plot.alpha.diversity, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap="Contexts ordered temporally from bottom to top based upon seriation results provided in the previous chapter. The middle two contexts (16RR1-BP2 and 16CD12-BP5) are interchangeable."}
library(dplyr)
library(ggplot2)
library(wesanderson)

pal <- wes_palette("Moonrise2", 13, type = "continuous")

gahagan_ct <- data[, -(1:2)]
labels <- paste0(data$context, " (", substr(data$region, 1, 1), ")")
rownames(gahagan_ct) <- labels

artifacts_long <- data.frame(context = rownames(gahagan_ct), 
                             stack(as.data.frame(gahagan_ct)))

artifacts <- rename(artifacts_long, Types = ind)

# configure plot
plot1 <- artifacts %>% 
  arrange(context) %>% 
  mutate(context = factor(context, levels = c(
                            "16CD12-BP1 (N)", "41CE19-F134 (S)", "16RR1-BP3 (S)",
                            "16CD12-BP5 (N)", "16RR1-BP2 (S)", "41CE19-F119 (S)", 
                            "16CD12-BP2 (N)", "16CD12-BP8 (N)"))) %>% 
  ggplot( aes(x=context, y = values, fill = Types)) + 
  geom_bar(stat="identity", position = "fill") + 
  coord_flip() +
  scale_fill_manual(values = pal) +
  labs(x = "Context",
       y = "Relative abundance (%)",
       title = "Abundance of diagnostic types across Gahagan-bearing contexts")

## print
plot1
```

## Diversity indices

```{r diversity.indices, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# Shannon diversity
## high diversity = more types and spread more evenly over types
data$H <- diversity(data[3:15])
H <- data$H
```

## Intra-site diversity

### Beta diversity - Mounds Plantation

```{r shannon.mp.context, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# filter dataset by mp context
# mutate to add temporal assignment per seriation
mp <- data %>% 
  filter(context == c("16CD12-BP1", "16CD12-BP2", 
                      "16CD12-BP5", "16CD12-BP8")) %>% 
  mutate(temporal = c(1, 3, 2, 4), .after = region)

pal2 <- wes_palette("Moonrise2", 4, type = "continuous")

# table of diagnostics
knitr::kable(mp, "pipe")

mp.H <- mp$H
mp.temporal <- mp$temporal

## plot
ggplot(mp, aes(x = mp.temporal, y = mp.H)) +
  geom_point(aes(shape = factor(context))) +
  geom_smooth(method = "lm")

# load
library(RRPP)

# are data normally distributed?
shapiro.test(mp.H)
hist(mp.H, col='steelblue')

# ANOVA: differences in assemblage diversity among the contexts?
mp.context <- mp$context

# MODEL: H as a function of context
mp.shannon <- lm.rrpp(mp.H ~ mp.context,
                      data = mp,
                      print.progress = FALSE,
                      iter = 9999)

# ANOVA: do contexts differ by H?
anova(mp.shannon)
```

### Beta diversity - George C. Davis

```{r shannon.gcd.context, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# filter dataset by mp context
# mutate to add temporal assignment per seriation
gcd <- data %>% 
  filter(context == c("41CE19-F134", "41CE19-F119")) %>% 
  mutate(temporal = c(1, 2), .after = region)

# table of diagnostics
knitr::kable(gcd, "pipe")

gcd.H <- gcd$H
gcd.temporal <- gcd$temporal

## plot
ggplot(gcd, aes(x = gcd.temporal, y = gcd.H)) +
  geom_point(aes(shape = factor(context))) +
  geom_smooth(method = "lm")

# load
library(RRPP)

# ANOVA: differences in assemblage diversity among the contexts?
gcd.context <- gcd$context

# MODEL: H as a function of context
gcd.shannon <- lm.rrpp(gcd.H ~ gcd.context,
                      data = gcd,
                      print.progress = FALSE,
                      iter = 9999)

# ANOVA: do contexts differ by H?
anova(gcd.shannon)
```

### Beta diversity - Gahagan Mound

```{r shannon.gm.context, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# filter dataset by mp context
# mutate to add temporal assignment per seriation
gm <- data %>% 
  filter(context == c("16RR1-BP2", "16RR1-BP3")) %>% 
  mutate(temporal = c(2, 1), .after = region)

# table of diagnostics
knitr::kable(gm, "pipe")

gm.H <- gm$H
gm.temporal <- gm$temporal

## plot
ggplot(gm, aes(x = gm.temporal, y = gm.H)) +
  geom_point(aes(shape = factor(context))) +
  geom_smooth(method = "lm")

# load
library(RRPP)

# ANOVA: differences in assemblage diversity among the contexts?
gm.context <- gm$context

# MODEL: H as a function of context
gm.shannon <- lm.rrpp(gm.H ~ gm.context,
                      data = gm,
                      print.progress = FALSE,
                      iter = 9999)

# ANOVA: do contexts differ by H?
anova(gm.shannon)
```

## Gamma diversity - Southern Behavioural Region Sites

## Gamma diversity - Combined

## Evenness

```{r even, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# evenness
## Pielou's J (Shannon diversity index divided by natural log of richness)
J <- H/log(S)
J
```

## Beta diversity

```{r diversity1.1, , out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap="Caddo burial assemblages that include Gahagan bifaces; 1, 41CE19-F134; 2, 41CE19-F119; 3, 16RR1-BP2; 4, 16RR1-BP3; 5, 16CD12-BP1; 6, 16CD12-BP2; 7, 16CD12-BP5; 8, 16CD12-BP8."}
data2 <- data[3:15]
types.jaccard <- vegdist(data2, method = "jaccard")
plot(
  hclust(types.jaccard),
  hang = -1,
  main = "Assemblages clustered by Jaccard similarity",
  axes = FALSE, ylab = ""
)
```

```{r diversity2, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap="Caddo burial assemblages that include Gahagan bifaces; 1, 41CE19-F134; 2, 41CE19-F119; 3, 16RR1-BP2; 4, 16RR1-BP3; 5, 16CD12-BP1; 6, 16CD12-BP2; 7, 16CD12-BP5; 8, 16CD12-BP8. The three sites in the lower part of the left quadrat exhibit lower diversity and evenness, and each represents the earliest context at each site to include a Gahagan biface. Those sites in the upper right quadrat have higher diversity and evenness, and occur later than those in the lower left quadrat."}
# summarize assemblage diversity to identify high & low diversity assemblages
library(maptools)
pch <- c(1, 3)[as.factor(data$region)]
plot(H, J, 
     pch = pch)
abline(h = median(J), 
       v = median(H), 
       lty = 2)
pointLabel(H, J, 
           rownames(data), 
           cex = .75)
leg.txt <- c(as.expression(bquote("Northern Behavioral Region")), 
             as.expression(bquote("Southern Behavioral Region")))
legend("bottomright", 
       leg.txt, 
       pch = c(1, 3))
```

```{r diversity3, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# sample size and richness
plot(S~N, 
     ylim = c(0, 12), 
     xlim = c(0, 184), 
     pch = 16)
abline(h = seq(0, 12, by = 2), 
       v = seq(0, 184, by = 50), 
       col = "black", 
       lty = 3)

# logarithmic function
data.log <- lm(S~log(N))
summary(data.log)
deviance(data.log)
xval <- seq(1, 184, by = 1)
lines(xval, 
      predict(data.log,
              data.frame(N = xval)),
      lty = 1)

# power function
data.pow <- lm(log(S)~log(N))
summary(data.pow)
sum((S-exp(fitted(data.pow)))^2)
lines(xval, 
      exp(predict(data.pow,
                  data.frame(N = xval))), 
      lty = 2)
```

```{r diversity4, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap="Rarefaction curve and confidence limits (dashed lines) for comparison of individual assemblages to the composite. Results demonstrate that diversity in the mortuary assemblages from 41CE19-F119 and 16RR1-BP3 are lower than expected based on the rarefaction curve."}
# rarefaction curve
## how do individual assemblages compare to the composite?
xval <- seq(2, 200, by = 2)
data.rar <- rarefy(T, xval, se = TRUE)
Est <- data.rar[1, ]
Sd <- data.rar[2, ]
rare <- cbind(lower = Est-2*Sd, 
              expected = Est, 
              upper = Est+2*Sd)

plot(S~N, 
     ylim = range(rare), 
     xlim = range(xval), 
     pch = 16)
matlines(xval, rare, 
         type = "l", 
         lty = c(2, 1, 2), 
         col = "black")
```
